---
title: 'Moderation analysis: insurance data'
author: "Jose M Sallan"
date: "23/5/2018"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Medical charges data

We have 1,338 observations of medical charges of patients in the US. The available variables are:

* **age**: age.
* **sex**: gender.
* **bmi**: body mass index.
* **children**: number of children.
* **smoker**: consumes tobacco regulary (yes, no).
* **region**: region where the patient lives.
* **charges**: medical charges (dependent variable).

Let's read the data:

```{r}
insurance <- read.csv("insurance.csv")
```

## Examining data

Let's examine relationships among variables plotting the dataset. We will obtain scatterplots for evey pair of variables. We will exclude gender and smoker, and we will plot in different colors smokers and non-smokers:

```{r}
plot(insurance[ , c(1,3,4,6,7)], pch=19, col=insurance$smoker)
```

There seems to be an interesting relationship between `charges` and `bmi` dependent on `smoker`:

## Interaction effect

Let's make a closeup of the `bmi` vs `charges` relationship:

```{r}
plot(insurance$bmi, insurance$charges, pch=19, col=insurance$smoker, xlab="bmi", ylab="charges")
```

The plot reveals that it seems to be that bmi influences medical charges (i.e., your health) only if you smoke. Let's examine the interaction effect:

```{r}
bmi.interaction <- lm(charges ~ bmi*smoker, data=insurance)
summary(bmi.interaction)
```

In the resulting model, the interaction term is significant, confirming analytically the moderating relationship. As the moderating variable is binary, we can analyze the `bmi` and `charges` relationship for both values of the variable:

```{r}
smokers <- insurance[which(insurance$smoker=="yes"), ]
nonsmokers <- insurance[which(insurance$smoker=="no"), ]

bmi.smokers <- lm(charges ~ bmi, data=smokers)
bmi.nonsmokers <- lm(charges ~ bmi, data=nonsmokers)

summary(bmi.smokers)
summary(bmi.nonsmokers)
```

Finally the regression coefficients of both models are significant, but the one of the smoker's model is larger than the one of the nonsmoker's. Let's plot both:

```{r}
plot(insurance$bmi, insurance$charges, col=insurance$smoker, xlab="bmi", ylab="charges")
abline(bmi.smokers, col="red", lwd=3)
abline(bmi.nonsmokers, col="black", lwd=3)
```

## Interaction effect with the rest of variables

A common practice when reporting interaction effects is presenting two models: one model with the variables of interest, and the other model with the variables and the interaction term. Then, comparing the fit of both models we can check if the addition of the interaction adds explanatory power to the model. Let's do it for that data:

```{r}
model1 <- lm(charges ~ age + sex + bmi + smoker, data=insurance)
model2 <- lm(charges ~ age + sex  + bmi*smoker, data=insurance)
library(stargazer)
stargazer(model1, model2, type="text")
```

Data seem to confirm that the interaction effect is significant. Let's assess it examining if model2 explains data better than model1:

```{r}
anova(model1, model2)
```

Data shows that we can discard the null hypothesis that both models are equivalent.

